---
title: "OBA - MPD analysis"
author: "Robyn L Ball, PhD"
date: "Dec 15, 2022"
output: 
  html_notebook:
    code_folding: hide
---

# PURPOSE

Calculate the coverage of OBA VT term mappings on MPD trait measures 

# METHODS

Using OBA VT term mappings, calculate coverage across MPD measures

# DATA

MPD ontology mappings and OBA-VT term mappings were downloaded on Dec 15, 2022.  
Put all data in /data/

MPD ontology mappings can be downloaded at https://phenomedoc-prod.jax.org/MPD_downloads/ontology_mappings.csv

OBA-VT mappings can be downloaded at https://github.com/obophenotype/bio-attribute-ontology/blob/master/src/mappings/oba-vt.sssom.tsv


# CODE

Set working directory to source code using Session -> Set working directory -> Source file location

If you do not have these packages, use install.packages("package_name") to install them or use Tools -> Install packages


```{r, message=FALSE, results='hide'}
options(stringsAsFactors = FALSE)
library(tidyverse)
library(reshape2)
library(DT)
```

read in data

```{r}
# OBA VT term mappings
oba <- read.delim("../data/oba-vt.sssom.tsv")
# MPD ontology terms to MPD measure mappings
terms <- read.delim("../data/ontology_mappings.csv", sep=",")
```

```{r}
print("OBA data preview")
head(oba)
```

```{r}
# VT terms in MPD
vts <- terms[grep("^VT", terms$ont_term), ]
```

distribution of OBA VT terms compared to MPD VT terms

```{r}
print(paste0("Number of VT terms mapped by OBA: ", length(unique(oba$subject_id))))
print(paste0("Number of VT terms with measures in MPD: ", length(unique(vts$ont_term))))
print(paste0("Number of VT terms mapped by OBA not in MPD: ", length(setdiff(oba$subject_id, vts$ont_term))))
print(paste0("Number of VT terms in MPD not mapped by OBA: ",length(setdiff(vts$ont_term, oba$subject_id))))
print(paste0("Number of VT terms mapped by OBA that are also in MPD: ", length(intersect(vts$ont_term, oba$subject_id))))
print(paste0("Percentage of VT terms in MPD that are also mapped by OBA: ", round(length(intersect(vts$ont_term, oba$subject_id))/length(unique(vts$ont_term))*100, 1)))
```
345 of 508 (67.9%) VT terms in MPD are mapped from OBA.  
163 VT terms in MPD are not mapped by OBA.

```{r}
meas <- unique(terms$measnum)
vt_meas <- unique(vts$measnum)
print(paste0("Number of MPD measures mapped to VT terms: ", length(vt_meas)))
print(paste0("Number of MPD measures: ", length(meas)))
print(paste0("Percentage of MPD measures mapped to VT terms: ", round(length(vt_meas)/length(meas)*100, 1)))

print(paste0("Number of MPD measures that can be mapped to from OBA: ",
             vts %>% 
  filter(ont_term %in% oba$subject_id) %>% 
  distinct(measnum) %>% 
  summarize(n=n())))
```

6325 of 6437 (98.3%) of MPD trait measures are mapped to VT terms.  
Of those mapped to VT terms, 5155 of 6325 (81.5%) MPD measures are mapped to VT terms also in OBA.

In total, OBA terms map to 80.1% of the trait measures in MPD. 


### Some context from other ontologies in MPD (MA, MP)

```{r}
terms %>% 
  filter(grepl("^MA", ont_term)) %>% 
  distinct(measnum) %>% 
  summarize(n=n(), p=n()/length(unique(terms$measnum)))

terms %>% 
  filter(grepl("^MP", ont_term)) %>% 
  distinct(measnum) %>% 
  summarize(n=n(), p=n()/length(unique(terms$measnum)))

terms %>% 
  add_column(ont=gsub(":.*", "", terms$ont_term)) %>% 
  select(measnum, ont) %>% 
  group_by(measnum) %>% 
  distinct() -> tb

tb %>% 
  summarize(n=n()) -> tb2

table(tb2$n)
table(tb$ont[which(tb$measnum %in% tb2$measnum[tb2$n==1])])
```
3935 (61.1%) MPD measures mapped to MA terms  

6273 (97.5%) MPD measures mapped to MP terms

Only 145 MPD measures are mapped to one ontology (46 to VT, 73 to MP, and 26 to MA).

## MPD measures mapped to VT terms not in OBA

163 VT terms not in OBA, sorted in order of how many MPD measures map to these terms:

```{r}
# vts %>% 
#   filter(!(ont_term %in% oba$subject_id)) %>% 
#   group_by(ont_term)

vts %>% 
  filter(!(ont_term %in% oba$subject_id)) %>% 
  group_by(ont_term, descrip) %>% 
  summarize(nmeas=n()) %>% 
  arrange(desc(nmeas))
```



