---
title: "oba-mpd-deeper-analysis"
author: "Robyn L Ball, PhD"
date: "Dec 15, 2022"
output: 
  html_notebook:
    code_folding: hide
---

# PURPOSE

Extract DOID to VT mappings using two methods:  
1. via OBA, HPO, and MONDO  
2. via DOID to gene to MP to MPD traits to VT terms  

# METHODS

Use existing ontologies to indirectly map these data.

# DATA

for (1), use  
+ http://purl.obolibrary.org/obo/mondo/mappings/mondo_exactmatch_doid.sssom.tsv  
+ http://purl.obolibrary.org/obo/mondo/mappings/mondo_exactmatch_omim.sssom.tsv  
+ https://github.com/monarch-initiative/mondo-ingest/releases/download/v2022-10-08/doid.sssom.tsv  
+ https://github.com/obophenotype/human-phenotype-ontology/releases/tag/v2022-12-15  
+ https://github.com/obophenotype/bio-attribute-ontology/blob/master/src/mappings/oba-vt.sssom.tsv  
+ https://api.triplydb.com/s/Um1dlK6_Y  

for (2), use  
+ https://download.alliancegenome.org/5.3.0/DISEASE-ALLIANCE/HUMAN/DISEASE-ALLIANCE_HUMAN_25.tsv.gz  
+ extract unique human genes and enter these in the MGI/MGD batch upload tool: https://www.informatics.jax.org/batch  
+ MPD ontology to trait measure mappings: https://phenomedoc-prod.jax.org/MPD_downloads/ontology_mappings.csv  

Data were downloaded on Dec 16, 2022 into /data/ directory

# CODE

Set working directory to source code using Session -> Set working directory -> Source file location

If you do not have these packages, use install.packages("package_name") to install them or use Tools -> Install packages

```{r}
options(stringsAsFactors = FALSE)
library(tidyverse)
library(reshape2)
```

read in data

```{r}
oba_vt <- read.delim("../data/oba-vt.sssom.tsv")
oba_hp <- read.csv("../data/OBA-HPO.csv")
oba_hp$hp <- gsub("http://purl.obolibrary.org/obo/", "", oba_hp$hp)
oba_hp$hp <- gsub("_", ":", oba_hp$hp)
oba_hp$oba <- gsub("http://purl.obolibrary.org/obo/", "", oba_hp$oba)
oba_hp$oba <- gsub("_", ":", oba_hp$oba)
mondo_omim <- read.delim("../data/mondo_exactmatch_omim.sssom.tsv", skip = 17)
mondo_doid <- read.delim("../data/mondo_exactmatch_doid.sssom.tsv", skip=16)
omim_hp <- read.delim("../data/phenotype.hpoa", skip=4)
terms <- read.delim("../data/ontology_mappings.csv", sep=",")
disease <- read.delim("../data/DISEASE-ALLIANCE_HUMAN_25.tsv", skip=15, head=T)
disease %>%
  select(DBObjectID, DBObjectSymbol, DOID, DOtermName) %>%
  distinct() -> dd
write.csv(unique(dd$DBObjectSymbol), "../data/disease_genes.csv", row.names = F, quote=F)
```

## Merge OBA data to get mappings

DOID -> MONDO -> OMIM -> HP -> OBA -> VT

DOID -> gene -> MP -> MPD measures -> VT

```{r}
oba_df <- mondo_doid[, c(1,2,4)]
colnames(oba_df) <- c("MONDO_id", "MONDO_label", "DOID")
# add OMIM
mondo_omim <- mondo_omim[, c(1,4)]
colnames(mondo_omim) <- c("MONDO_id", "OMIM_id")
oba_df <- merge(oba_df, mondo_omim, by="MONDO_id", all.x=T, all.y=F, sort=F)
# add HP
omim_hp <- omim_hp[, c(1,2,4)]
colnames(omim_hp) <- c("OMIM_id", "OMIM_label", "HP_id")
oba_df <- merge(oba_df, omim_hp, by="OMIM_id", all.x=T, all.y=F, sort=F)
# merge with OBA
colnames(oba_hp) <- c("HP_id", "HP_label", "OBA_id", "OBA_label")
oba_df <- merge(oba_df, oba_hp, by="HP_id", all.x=T, all.y=F, sort=F)
# merge with VT
oba_vt <- oba_vt[, c(1,3)]
colnames(oba_vt) <- c("VT_id", "OBA_id")
oba_df <- merge(oba_df, oba_vt, by="OBA_id", all.x=T, all.y=F, sort=F)
write.csv(oba_df, "oba-hp-omim-mondo-do-vt.csv", row.names = F)
oba_df$has_VT <- as.numeric(!is.na(oba_df$VT_id))
table(oba_df$has_VT)
```

## DOID -> human gene -> MP -> mouse traits -> VT 

Uploaded to MGI batch upload tool: http://www.informatics.jax.org/batch

Got MP terms associated with these genes: disease_genes_MGIBatchReport_20221215_162329.txt

```{r}
mp <- read.delim("../data/disease_genes_MGIBatchReport_20221215_162329.txt")
```

for each disease (DOID), find MP terms associated with disease-related genes.

Use these MP terms in MPD to find traits and associated VT terms

```{r}
mapping <- data.frame()
doids <- unique(dd$DOID)
length(doids)
for (doid in doids) {
  genes <- unique(dd$DBObjectSymbol[which(dd$DOID==doid)])
  mps <- unique(mp$MP.ID[which(mp$Input %in% genes)])
  traits <- unique(terms$measnum[terms$ont_term %in% mps])
  vterms <- unique(terms$ont_term[terms$measnum %in% traits])
  vterms <- vterms[grep("^VT", vterms)]
  terms %>%
    filter(ont_term %in% vterms) %>%
    group_by(ont_term) %>%
    distinct() %>%
    summarize(nmeas=n()) -> mm
  mapping <- rbind(mapping, data.frame(DOID=rep(doid, nrow(mm)), mm))
}
write.csv(mapping, "../data/DOID2gene2MP2VT_mapping.csv", row.names = F)
```


