---
title: "OBA DOID-VT comparison"
author: "Robyn L Ball, PhD"
date: "Dec 19, 2022"
output: 
  html_notebook:
    code_folding: hide
---

# PURPOSE

Compare DOID-VT mappings we find through OBA and other ontologies vs DOID-VT mappings we find through MP phenotypic alleles and MPD trait measures.

# METHODS

### Approach 1

DOID -> human gene (via AGR disease map) -> MP terms (via MGI, phenotypic allele) -> mouse trait measures (MPD)  -> VT terms (via MPD mouse trait measures)

### Approach 2

DOID -> MONDO -> OMIM -> HP -> OBA -> VT

### Comparison

1. How many DOID-VT mappings do we get from each approach?  
2. Of the DOID-VT OBA mappings, restrict the VT terms to those in MPD. How much overlap do we find between the approaches?

# RESULTS

Using both approaches, we identified 4910 disease terms associated with VT terms. Both approaches identified VT terms for the same 1348 DOIDs. Of these 1348 disease terms, 558 (41.4%) had associated VT terms with some overlap (mean overlap of VT terms was 26.6% for OBA and 16.7% for MP-VT).  


OBA mapped to 3416 of 3455 (98.9%) DOIDs mappable through HP.  
Of these 3416 DOIDs, 3408 (99.8%) mapped to VT terms via OBA.  
Of these 3408 DOIDs, 3033 (89%) mapped to VT terms with mouse trait measures in MPD.  
Thus, in total, OBA mapped 3033 of 3455 (87.8%) DOIDs to VT terms with mouse trait measures in MPD.

The union of DOID terms retrieved by both approaches contains 5457 DOIDs associated with VT terms.  
1872 of these 5457 DOIDs were mapped to VT terms with both approaches.    
Of these 1872 DOIDs, 799 (42.7%) mapped to at least one of the same VT terms.  

3033 DOIDs annotated to VT terms by OBA semantic mappings comprise 51.9% of trait measures in MPD annotated to VT terms

See https://docs.google.com/document/d/1fkXIvngeN8Kv3WL4Yr9E3aKj9WXv3wAaIoN4GC-GIto/edit?usp=sharing


# CODE

```{r}
require(tidyverse)
require(reshape2)
options(stringsAsFactors = FALSE)
```

read in DOID-VT mappings

```{r}
mp_vt <- read.csv("../data/DOID2gene2MP2VT_mapping.csv")
oba_df <- read.csv("../data/oba-hp-omim-mondo-do-vt.csv")
terms <- read.delim("../data/ontology_mappings.csv", sep=",")
disease <- read.delim("../data/DISEASE-ALLIANCE_HUMAN_25.tsv", skip=15)
```

## Summary

```{r}
n_d <- length(unique(oba_df$DOID[which(!is.na(oba_df$HP_id) & !is.na(oba_df$DOID))]))
n_o <- length(unique(oba_df$DOID[which(!is.na(oba_df$OBA_id))]))
print(paste0("OBA mapped to ", n_o, " of ", n_d," (", round(n_o/n_d*100, 1), "%) DOIDs mappable through HP."))
n_v <- length(unique(oba_df$DOID[which(grepl("^VT", oba_df$VT_id) & !is.na(oba_df$HP_id) & !is.na(oba_df$DOID))]))
print(paste0("Of these ", n_o, " DOIDs, ", n_v, " (", round(n_v/n_o*100, 1), "%) mapped to VT terms via OBA."))
vt_mpd <- unique(terms$ont_term[grep("^VT", terms$ont_term)])
n_v_mpd <- length(unique(oba_df$DOID[which(oba_df$VT_id %in% vt_mpd)]))
print(paste0("Of these ", n_v, " DOIDs, ", n_v_mpd, " (", round(n_v_mpd/n_v*100, 1), "%) mapped to VT terms with mouse trait measures in MPD."))
print(paste0("Thus, in total, OBA mapped ", n_v_mpd, " of ", n_d, " (", round(n_v_mpd/n_d*100,1), "%) DOIDs to VT terms with mouse trait measures in MPD."))
```

### OBA approach summary

OBA mapped to 3416 of 3455 (98.9%) DOIDs mappable through HP.  
Of these 3416 DOIDs, 3408 (99.8%) mapped to VT terms via OBA.  
Of these 3408 DOIDs, 3033 (89%) mapped to VT terms with mouse trait measures in MPD.  
Thus, in total, OBA mapped 3033 of 3455 (87.8%) DOIDs to VT terms with mouse trait measures in MPD.


```{r}
m_d <- length(unique(mp_vt$DOID))
print(paste0(m_d, " DOIDs were mapped to VT terms via MP and MPD trait measures."))
m_inter <- length(intersect(mp_vt$DOID, oba_df$DOID[which(!is.na(oba_df$VT_id))]))
print(paste0("Of ", m_d, " DOIDs, ", m_inter, " (", round(m_inter/m_d*100, 1), "%) were in common with DOIDs mapped to VT terms through OBA."))
```


### MP-VT approach summary

Though phenotypic alleles and mouse trait measure mappings in MPD, 3225 DOIDs were mapped to VT terms.  
Of these 3225 DOIDs, 1474 (45.7%) were in common with DOIDs mapped to VT terms through OBA.  

### Pathway 

```{r}
print(paste0("Of ", length(unique(oba_df$DOID)), " DOIDs mapped to MONDO, ", length(unique(oba_df$DOID[!is.na(oba_df$OMIM_id)])), " had OMIM mappings and ", length(unique(oba_df$DOID[!is.na(oba_df$HP_id)])), " had OMIM-HP mappings."))
```

Of 9815 DOIDs mapped to MONDO, 3642 had OMIM mappings and 3455 had OMIM-HP mappings.

```{r}
print(paste0("Of ", length(unique(disease$DOID)), " annotated to human genes, ", length(unique(mp_vt$DOID)), " mapped to MP terms and to VT terms through MPD trait measures."))
print(paste0("Of these ", length(unique(disease$DOID)), " DOIDs annotated to human genes, ", length(intersect(oba_df$DOID[!is.na(oba_df$VT_id)], disease$DOID)), " were also mapped to VT terms via OBA."))
```

Of 5835 DOIDs annotated to human genes in AGR, 3225 mapped to MP & VT terms through MPD trait measures.  
And, of these 5835 DOIDs annotated to human genes, 3048 were also mapped to VT terms via OBA.  

### In common

Consider only those OBA mappings that also mapped to VT terms in MPD

```{r}
doids <- union(mp_vt$DOID, oba_df$DOID[which(oba_df$VT_id %in% vt_mpd)])
df <- data.frame()
for (id in doids) {
  vt1 <- unique(mp_vt$ont_term[which(mp_vt$DOID==id)])
  n1 <- length(vt1)
  vt2 <- unique(oba_df$VT_id[which(oba_df$DOID==id & !is.na(oba_df$VT_id))])
  n2 <- length(vt2)
  vt3 <- intersect(vt2, terms$ont_term)
  n3 <- length(vt3)
  ncommon <- length(intersect(vt1, vt2))
  ncommon_mpd <- length(intersect(vt1, vt3))
  p1 <- p2 <- p3 <- 0
  if (ncommon > 0) {
      p1 <- round(ncommon/n1*100, 1)
      p2 <- round(ncommon/n2*100, 1)
      p3 <- round(ncommon_mpd/n3*100, 1)
  }
  df <- rbind(df, data.frame(DOID=id, ncommon_MPD=ncommon_mpd, ncommon=ncommon, num_1=n1, num_2=n2, num_3=n3, per_1=p1, per_2=p2, per_3=p3))
}
```

```{r}
dd <- df[, c(1,2,4,6,7,9)]
colnames(dd) <- c("DOID", "ncommon_VT", "num_VT_MP", "num_VT_OBA", "percent_common_VT_MP", "percent_common_VT_OBA")
dd <- merge(dd, disease[, c("DOID", "DOtermName")], all.x=T, all.y=F, sort=F)
dd <- unique(dd)
summary(dd[, -c(1,7)])
```

```{r}
any(is.na(dd$DOID))
print(paste0("Using both approaches, we identified VT terms for ", nrow(dd), " DOIDs."))
both <- which(dd$DOID %in% oba_df$DOID[which(oba_df$VT_id %in% vt_mpd)] & dd$DOID %in% mp_vt$DOID)
print(paste0(length(both), " DOIDs were mapped to VT terms with both approaches."))
print(paste0("Of these ", length(both), " DOIDs, ", length(which(dd$ncommon_VT > 0)), " (", round(length(which(dd$ncommon_VT > 0))/length(both)*100,1), "%) mapped to at least one of the same VT terms."))

ds <- dd[intersect(which(dd$num_VT_OBA > 0), both), ]
print(paste0("Of these ", length(both), " DOIDs, ", length(which(ds$ncommon_VT > 0)), " (", round(length(which(ds$ncommon_VT > 0))/length(both)*100,1), "%) mapped to at least one of the same VT terms."))
```

Using both approaches, we identified VT terms for 5457 DOIDs.  
1872 of these 5457 DOIDs were mapped to VT terms with both approaches.  
Of these 1872 DOIDs, 799 (42.7%) mapped to at least one of the same VT terms.


```{r}
ds %>% 
  arrange(desc(ncommon_VT), desc(percent_common_VT_OBA), desc(percent_common_VT_MP))
```
## Coverage in MPD

```{r}
doids <- unique(oba_df$DOID[which(oba_df$VT_id %in% vt_mpd)])
vt_meas <- unique(terms$measnum[grep("^VT", terms$ont_term)])
nmeas <- length(vt_meas)
n <- length(unique(terms$measnum))
oba_meas <- unique(terms$measnum[which(terms$ont_term %in% oba_df$VT_id)])
n_oba <- length(oba_meas)
print(paste0(length(doids), " DOIDs annotated to VT terms by OBA semantic mappings comprise ", round(n_oba/nmeas*100, 1), "% of trait measures in MPD annotated to VT terms"))
```

